---
title: "Exploration of Petroleum Rock Samples"
output:
  html_document: default
  html_notebook: default
---

*Adam McCarthy*

Exploratory data analysis of rock core barrel scans and their permeability. 

Data-set consists of four variables.
Of these permeability could be thought of as a dependent variable.
While the other variables are assumed to be independent.

* area = area of pores space, in pixels out of 256 by 256
*	peri = perimeter in pixels
*	shape = perimeter/sqrt(area)
*	perm = permeability in milli-Darcies

There is little information about these variables
and data-set.

Either this was from microscopy or CT scans of the
cross-sections of core.

The source information is:

> Data from BP Research, image analysis by Ronit Katz, U. Oxford.

The details state that information comes from 12 core samples
with 4 cross-sections taken.

4 cross sections from 12 cores give 48 samples.

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load dependancies
library(ggplot2)
library(ggthemes)
library(scales)
library(extRemes)
library(sm)
library(gridExtra)
library(GGally)
```

```{r echo=FALSE}
# Get data
help(rock)
data(rock)

# Create a new categorical variable for quality
rock['perm_c'] <- ifelse(rock$perm <= 300, "low", "permeabale")
```

#### Descriptive Statistics
```{r echo=FALSE}
# Descriptive Statistics
summary(rock)
```

The summary of the data-set highlights the distribution of the data. 
The data consist
of four original variables and one categorical variable created based on
a permeability cut-off of 300mD (milli-Darcies). The categorical variable
is to highlight the difference between relatively poor ("low") and better
("permeable") reservoir.

All original variables are numeric.

Area of pore space is measured out of pixels of 256 by 256. This gives
a total number of pixels of 65536. So the maximum value has pore space
of ~19% compared to non-pore space. Following this approach
a new variable is calculated to give area as a percentage as this is
easier to comprehend.

The perimeter would count the geometric boundary of the
pore space in pixels. There is no information present to convert
this into a measurement.

Shape would reflect some information about the shape of the pore
space. No information is at hand other than it is perimeter/sqrt(area)

Permeability has very low values at it´s minimum of 6.3 mD up to a max
of 1300mD.

#### Checking data
```{r echo=FALSE, new_var, fig.height = 5, fig.width = 5}
# Area percentage by converting total number of pixels
# into the relative number of pixels compared to
# 256*256.
rock['area_perc'] <- (rock$area/65536)
```

```{r echo=FALSE, var_check, fig.height = 5, fig.width = 5}
ggplot(aes(x = rock$shape, y = (rock$peri/sqrt(rock$area))),
       data = rock) +
  geom_point(alpha = 0.2,
             position = position_jitter(),
             color = '#192231') +
  theme_tufte() +
  labs(
    title = "Check on shape",
    x = "Shape perimeter/sqrt(area)",
    y = "Variables (perimeter/sqrt(area))"
    ) +
  geom_smooth(method = 'lm', formula = y~x,
              colour = '#192231')
```

The first plot is made to check what the shape variable is based on.
In the notes it is described as perimeter/sqrt(area) but when plotted against
this axis they do not correlate. If the shape variable was a derivative of
area and perimeter then this would be a matching along a straight line.

This would imply that shape has been calculated on each individual pore
rather than the sum for each cross-section, which is what is at hand.

The ratio would suggest that for the same perimeter larger pores would
have lower values. While if the perimeter was larger for the same pore
area the value would be larger.

Circular pores with an even shape would have larger pore area compared
to perimeter so would have lower values. While highly irregular geometric
shapes would have a longer perimeter for their area given higher values.

The tricky thing here is we are dealing with the sum of these
values for the cross section. So this is an aggregate of all these values.

#### Uni-varaite analysis
```{r echo=FALSE, Multiplot_function}
# Entirely taken from:
# http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/
# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```

```{r echo=FALSE, U2, fig.height = 3, fig.width = 5, fig.align="center"}
small_hist <- function(data_v, name, scale, df=rock) {
plt <- ggplot(data = df, aes(data_v)) + 
  geom_histogram(bins = 20,
                 col = "white",
                 fill = "#9099A2"
                 ) +
  theme_tufte() +
  labs(
    title = name
    ) +
  labs(
    x = scale,
    y = "Count"
    )

return(plt)
}

plt1 <- small_hist(rock$area_perc, "Area of pore space", "%")
plt2 <- small_hist(rock$peri, "Perimeter", "Pixels")
plt3 <- small_hist(rock$shape, "Shape", "pores perimeter/sqrt(area)")
plt4 <- small_hist(rock$perm, "Permeability", "milli-Darcies")
multiplot(plt1, plt2, plt3, plt4, cols = 2)
```

Histogram better represent the distribution of values. Given 48 samples only 20 bins have been
used. 

The Area of pore space shows an apparent normal distribution.

Shape shows a skewed distribution.

Perimeter shows a bi modal distribution.

Permeability shows a large group of values close to 0 and then
groups of bars above 500mD. A cut-off of 300mD has been used as
an arbitary number to seperate between low and high values.

```{r echo=FALSE, QQ_2, fig.height = 5, fig.width = 5, fig.align="center"}
qqnorm(rock$peri) 
qqline(rock$peri)
```

Checking perimeter in a QQ plot highlights the bi modal distribution.
Given the sparse number of samples the confidence bands highlight that
it may still be sourced from a normal distribution.

```{r echo=FALSE, QQ_3, fig.height = 5, fig.width = 5, fig.align="center"}
qqnorm(rock$area_perc)
qqline(rock$area_perc)
```

Area of pore space does show a normal distribution, one with a number of clusters
in this data-set.

```{r echo=FALSE, QQ_4, fig.height = 5, fig.width = 5, fig.align="center"}
qqnorm(rock$shape)
qqline(rock$shape)
```

Shape shows it skews to the right, it intersects the confidence interval
suggesting that even with this number of data points it is unlikely 
it is sourced from a normal distribution.

All of these observations, however, should be treated with caution, we are
dealing with 12 core samples, perhaps from a very similar setting. This
does not mean these observations may be relevant for a broader sample or the
population of values.

#### Bi-variate analysis
```{r echo=FALSE, BPlots, fig.height = 5, fig.width = 4, fig.align="center"}
freq_poly <-  function(x_v, name, scale, 
                       dataset=rock, group=rock$perm_c) {
plt <- qplot(x = x_v, data = dataset,
             geom = 'freqpoly', 
             color = group,
             bins = 15) +
  scale_x_continuous() +
  theme_tufte() +
  labs(
    title = name
    ) +
  labs(
    x = scale
      )
plt + scale_color_manual(values = c('#0B3C5D', '#D9B310'))
}

pb1 <- freq_poly(rock$area_perc, "Area of pore space", "%")
pb2 <- freq_poly(rock$peri, "Total perimeter of pores", "pixels")
pb3 <- freq_poly(rock$shape, "Shape", "Pores perimiter/sqrt(area)")

multiplot(pb1, pb2, pb3, cols = 1)
```

Using the categorical variable separating permeability each variable shows
some relationship to permeability.

The first plot shows spikes where there are fewer permeable samples with
an increase in pore area. This in counter-intuitive as an initial perspective
would be more pore space would lead to more permeability.

The second plot suggests that the total length of pore perimeter has a
strong relationship to the two bi modal peaks of perimeter. In which permeable
reservoirs have a lower total perimeter.

The final plot showing the shape variable gives lower values 
as having low permeability.
These would be the more uniform circular geometries. More irregular geometries
would give some of the permeable results but there are also low values here
as well.

```{r echo=FALSE, freq_pair_plot, fig.height = 9, fig.width = 9, message=FALSE, warning=FALSE,}
ggpairs(rock,
        title = "Rock sample pair plot"
        ) +
  theme_minimal()
```

The most correlated variables are area and perimeter (0.82).

Permeability shows correlations to all three variables with -0.74
to perimeter being the strongest.

```{r echo=FALSE, scatter_den, fig.height = 5, fig.width = 5}
ggplot(aes(x = rock$area_perc, y = rock$peri),
       data = rock) +
  geom_point(alpha = 0.5,
             color = '#192231') +
  theme_tufte() +
  labs(
    title = "Area vs Perimeter",
    x = "%",
    y = "pixels"
    ) +
  geom_smooth(method = 'lm', formula = y~x,
              colour = '#192231')
```

Area compared to perimeter has the highest correlation of 0.82, the plot
of these two variables highlights this falls into two clusters.
Overall it suggests a larger area has a larger overall perimeter length
which would make sense.

```{r echo=FALSE, scatter_den2, fig.height = 5, fig.width = 5}
ggplot(aes(x = rock$area_perc, y = rock$shape),
       data = rock) +
  geom_point(alpha = 0.5,
             color = '#192231') +
  theme_tufte() +
  labs(
    title = "Area vs Shape of Pore Space",
    x = "%",
    y = "Pores perimiter/sqrt(area)"
    ) +
  geom_smooth(method = 'lm', formula = y~x,
              colour = '#192231')
```

Area compared to shape does not show any clear correlations. Different
sums for pore shape geometry can occur for different total areas.

```{r echo=FALSE, scatter_den3, fig.height = 5, fig.width = 5}
ggplot(aes(x = rock$peri, y = rock$shape),
       data = rock) +
  geom_point(alpha = 0.5,
             color = '#192231') +
  theme_tufte() +
  labs(
    title = "Perimieter vs Shape of Pore Space",
    x = "pixels",
    y = "Perimiter/sqrt(area)"
    ) +
  geom_smooth(method = 'lm', formula = y~x,
              colour = '#192231')
```

Perimeter of pore space compared to the shape gives two clusters along
the x axis (perimeter) each having a range of shapes.

```{r echo=FALSE, scatter_den4, fig.height = 5, fig.width = 5}
ggplot(aes(x = rock$peri, y = factor(rock$perm)),
       data = rock) +
  geom_point(alpha = 0.8,
             color = '#192231') +
  theme_tufte() +
  labs(
    title = "Perimieter vs Permeability",
    x = "pixels",
    y = "milli-Darcies"
    )
```

Starting to compare the variables to the dependent permeability perimeter
had the highest correlation. The two clusters in perimeter appear to
relate to to different trends in permeability. With lower perimeter lengths
relating to higher permeability.

```{r echo=FALSE, scatter_den5, fig.height = 5, fig.width = 5}
ggplot(aes(x = rock$shape, y = factor(rock$perm)),
       data = rock) +
  geom_point(alpha = 0.8,
             color = '#192231') +
  theme_tufte() +
  labs(
    title = "Pore shape vs Permeability",
    x = "Pores perimiter/sqrt(area)",
    y = "milli-Darcies"
    )
```

Pore shape compared to permeability suggest that the highest values of
pore shape are often associated to higher permeability with one outlier.
There is a lot of overlap in pore shape values and the range of permeability.

```{r echo=FALSE, scatter_den6, fig.height = 5, fig.width = 5}
ggplot(aes(x = rock$area_perc, y = factor(rock$perm)),
       data = rock) +
  geom_point(alpha = 0.8,
             color = '#192231') +
  theme_tufte() +
  labs(
    title = "Area of pore space vs Permeability",
    x = "%",
    y = "milli-Darcies"
    )
```

Area of pore space shows a confusing plot with each value of permeability
having a range of pore areas. Often highest pore area has lowest permeability.

#### Multi-variate analysis
```{r echo=FALSE, Multivariate_Plots_7, fig.height = 5, fig.width = 8}
plt_mv7 <- ggplot(aes(
  x = (rock$area_perc/rock$shape),
  y = (rock$peri)
  ),
       data = rock) +
  geom_point(alpha = 0.01,
             position = position_jitter()
             ) +
  theme_tufte() +
  labs(
    title = "Attempting to seperate permeability groups",
    x = "Area % / Shape",
    y = "Perimiter (pixels)"
    )
fnl_plt_4 <- plt_mv7 + geom_point(aes(size = (perm)), 
                                  pch = 21, 
                                  show.legend = T,
                                  alpha = 0.6) + 
  aes(fill = factor(perm_c))
fnl_plt_4
```

After struggling to find strong relationships between individual variables and
permeability this leads to multi-variate analysis. Using two variables as
ratios compared to a third variable will allow to explore if it is possible
to separate out different permeability groups. As perimeter has the
strongest relationship to permeability this is kept on one axis while
area and shape are combined into a ratio for the x axis.

The figure shows the best visual representation of separating the low and
permeable intervals. This plot suggests, for this data-set, that relatively
lower perimeter values strongly separate permeable reservoir units.

The combination of pore area and pore shape further help separate out this
trend. Lower values of this ratio equate to lower pore area and more irregular
pore geometries.

This is the trends apparent in this data-set, however, these observations
are counter intuitive. The greater the pore area, with longer pore perimeters
would be associated with being easier for fluids to flow through.
Less interfacial tension between grains and thus higher
permeability values.

It could relate to micro vs. macro pore space as permeability is controlled
by macro pore space rather than micro. Given that all these values are sums
for the cross section if micro pore spaces are included this could be creating
the greater area values and greater perimeter values. Furthermore, the
irregular pore space being associated with more permeable layers may reflect
this trend with micro-pores being typically more uniform. So if the sample
is dominated by macro pores, perhaps this is reflected in the sum of the
pore shape values.

This could be a reflection of local conditions from where these cores are
taken from. For example carbonate compared to siliciclastic depositional
settings.

Diagenesis, chemical alterations during burial of sediments can lead to fibrous
growths (e.g. illite cement) between pore spaces which can have dramatic
effects on permeability while retaining large pore spaces.

Without further information about the local conditions of each core sample
it is not possible to take this assessment any further. The observations
presented here should not be generalized to other similar data-sets.

```{r echo=FALSE, Multivariate_Plots_8, fig.height = 6, fig.width = 9}
plt_mv8 <- ggplot(aes(
  x = (rock$area_perc/rock$shape),
  y = (rock$peri)),
       data = rock) +
  geom_point(alpha = 0.1,
             position = position_jitter()
             ) +
  theme_minimal() +
  labs(
    title = "Comparing key variables to permeability",
    x = "Area/Shape",
    y = "Perimeter"
    )
plt_mv8 <- plt_mv8  + facet_wrap(~ factor(perm))
plt_mv8 <- plt_mv8 + geom_point(aes(size = log10(perm)), 
                     pch = 21,
                     show.legend = T) + 
  aes(fill = factor(perm), alpha = 0.6)
plt_mv8
```

The final plot separates out all of the permeability values, this also
highlights that there are only 12 measurements of permeability out of all
48 samples. This reflects that each core likely only has one permeability
measurement. So even though there are 48 samples there are only 12 samples
of permeability. Based on this assumption the above plot shows each
core and their four values.

Each of these plots shows a certain grouping but there is variance in the
values of each cross-section within each core. There is less variation
in perimeter (y axis) compared to the range in the x axis (Area/Shape).

A few of the cores break this observation, for example 950mD has a range of
perimeter values.

From 142mD up to 1300mD the change in perimeter values can be seen. 
Area/shape has less of a consistent trend, i.e. as permeability increases
there is not a consistent relationship. This would suggest it is not possible
to take this analysis further than separating permeable from lower permeability
samples. If the threshold for this categorical value is changed to < 140mD 
instead of 300mD the relationship would be stronger.